---
title: "Linear Regression - Practice 1"
author: "Debbie Cannon"
date: "01/03/2022"
output: html_document
---

Repeat this analysis for different predictors, or a different outcome. Remember that if you use categorical predictors, you will need to convert them to numeric first.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
library(tidyverse)    # for general data science
library(tidymodels)   # for machine learning
library(corrplot)     # for visualising correlation matrices
library(vip)          # for variable importance plots

```

```{r read in data}
scot_data <- read_csv("Scotland-Vulnerability-Resource_v0.2.csv")

glimpse(scot_data)
```

```{r remove the postcode}
DZ_data <- scot_data %>%
  select(-Postcode) %>%
  drop_na() %>%    # remove rows with missing values
  unique()         # keep unique rows
```

```{r remove geography names and population numbers}
DZ_data <- DZ_data %>%
  select(-Data_Zone,-Intermediate_Zone,-Council_area,-NHS_Health_Board_Region,-Total_population,-Working_age_population,-URname)
```

Let's see if we can predict the variable "DRUG".

```{r}
DZ_data %>%
ggplot(aes(x = DRUG)) + 
  geom_histogram(bins = 30, col= "white")
```

#need to transform the data because it is skewed to the right.

I am going to use the variable Standard Mortality Rate (SMR) instead.


```{r checking to see if SMR is a normal distribution}
DZ_data %>%
ggplot(aes(x = SMR)) + 
  geom_histogram(bins = 30, col= "white")
```
I will try using SMR as this is less skewed than the DRUG variable.
```{r data-splitting}

# Create the split
DZ_split <- DZ_data %>%
    initial_split(prop = 0.8,        # proportion in training set
                  strata = SMR)  # ensure distribution of SMR in the two datasets in similar

# Save the two datasets as objects
DZ_train <- training(DZ_split)
DZ_test <- testing(DZ_split)

```

```{r data-corr}

DZ_cor <- cor(DZ_train %>% select_if(is.numeric))  # we can only use numeric variables to calculate correlation
corrplot(DZ_cor, tl.cex = 0.5)

```
```{r scatterplot income_rate}

DZ_train %>%
  ggplot(aes(x = Income_rate, y = SMR)) + 
  geom_point()

```

```{r scatterplot Employment_rate}

DZ_train %>%
  ggplot(aes(x = Employment_rate, y = SMR)) + 
  geom_point()

```

```{r scatterplot CIF (Comparative illness factor)}

DZ_train %>%
  ggplot(aes(x = CIF, y = SMR)) + 
  geom_point()

```
```{r scatterplot EMERG Emergency stays in hospital}

DZ_train %>%
  ggplot(aes(x = EMERG, y = SMR)) + 
  geom_point()
```


```{r specify-recipe}

simple_rec <- DZ_train %>%
  recipe(SMR ~ Income_rate + Employment_rate + CIF + EMERG) %>%
  step_zv(all_predictors()) %>%
  step_corr(all_predictors())
```

```{r specify-model}

## A linear regression model specification with default settings
lm_spec <- linear_reg() 

```


```{r specify-workflow}

DZ_wflow <-workflow() %>%
           add_recipe(simple_rec) %>%
           add_model(lm_spec)

```

```{r train-model}

DZ_wflow_fit <- DZ_wflow %>%
    fit(data = DZ_train)

```


```{r extract-model}

fitted_model <- extract_fit_engine(DZ_wflow_fit)

```

```{r residuals-fitted}
plot(fitted_model, 1)
```

```{r q-q}
plot(fitted_model, 2)
```


```{r scale-location}
plot(fitted_model, 3)
```

```{r variable-importance}

DZ_wflow_fit %>% 
  extract_fit_parsnip() %>% 
  vip(num_features = 2)

```


```{r eval-testing-data}

# Use the model for prediction on the testing data now
predicted_SMR_test <- DZ_wflow_fit %>%
  predict(new_data = DZ_test)

# As before, add the predicted column to the dataset so we have everything in one dataframe
results_test <- DZ_test %>%
  bind_cols(predicted_SMR_test) %>%
  rename(pred_SMR = .pred)   # rename the predicted column

# Evaluate the performance
metrics(results_test, truth = SMR, estimate = pred_SMR)

```

```{r plot-model-prediction}

results_test %>%
  ggplot(aes(x = SMR, y = pred_SMR)) +
  geom_abline(intercept=0, slope=1) +  # we want data to fall close to this line
  geom_point() +
  xlab("Actual outcome values") + 
  ylab("Predicted outcome values")

```








